---
title: "Homework 4"
author: "Fuyu"
date: "4/9/2018"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Q1

## Part a: Model this outcome as a function of treatment assignment using a Poisson regression. Does the model fit well? Is there evidence of overdispersion?

The model does not fit well, since the deviance ratio is very large $\frac{12925}{431} = 29.9884 $, though the coefficient are all statistically significant. 

The diviance decreases from 13299 in null model to 12925 in model 1 is significant, which means the intervention changes the outcome. In treatment group 1 (couple = 1, women_alone = 0), only women of couples participated in the counseling sessions. Couples in this group have about a 32.24% lower chance than group where both members of couples participated in the session of having unprotected sexual behaviors. In treatment group 2 (couples = 0, women_alone = 1), women who participated the study alone have 57.212% lower chance to have unprotected sexual behavior compared to group where both members of couples participated in the session.

```{r}
library('data.table')
library('knitr')

df = fread('risky_behaviors.csv')

df$treatment <- c("0")
df$couples <- as.numeric(df$couples)
df$women_alone <- as.numeric(df$women_alone)
df <- data.frame(df)
df$fupacts <- round(df$fupacts, digits = 0)

for (i in 1:nrow(df)){
  if ((df[i,3] == "0") == TRUE &
      (df[i,4] == '0') == TRUE) {
    df[i,8] = "0"
} else if ((df[i,3] == "1") == TRUE &
      (df[i,4] == '0') == TRUE){
    df[i,8] = "1"
} else if ((df[i,3] == "0") == TRUE & 
      (df[i,4] == '1') == TRUE){
    df[i,8] = "2"
}
}

data = df
data$treatment = factor(data$treatment)

Mod1 = glm(fupacts ~ treatment, data = data, family = 'poisson')
summary(Mod1)

```

And there is overdispersion. The sum of squared standard residuals is  $\sum_{z = 1}^{m}{z_i} = 19022$ , compared to an expected value of n - k = 431. The estimated overdispersion factor is $\frac{\sum_{i = 1}^{m}{z_i}}{n-k} = \frac{19022}{431} = 44.13458$, and the p-value is 1, indicating that the probability is essentially zero that a random variable from a $\chi^2_{431}$ distribution would be as large as 19022. In summary, the risk_behaviors data is overdispersed by a factor of 44, which is huge. 

```{r}
pv = fitted(Mod1)
r = (df$fupacts - fitted(Mod1))

library(ggplot2)
library("gridExtra")
p1 = ggplot() +
  geom_point(aes(pv,r),colour = "blue", size = 1,show.legend = TRUE) +
  ylab('raw residuals') + 
  xlab('predicted value') +
  geom_hline(yintercept=0, color = "red") +
  theme_bw()

sr <- (df$fupacts - fitted(Mod1))/sqrt(fitted(Mod1))

p2 = ggplot() +
  geom_point(aes(pv,sr),colour = "blue", size = 1,show.legend = TRUE) +
  ylab("standardized residuals") + 
  xlab("predicted value") +
  geom_hline(yintercept=0, color = "red") +
  geom_hline(yintercept=1.96, linetype="dashed", color = "red") +
  geom_hline(yintercept=1.96, linetype="dashed", color = "red") +
  geom_hline(yintercept=-1.96, linetype="dashed", color = "red") +
  theme_bw()
grid.arrange(p1,p2)

yhat = predict(Mod1,type="response")
sum(sr^2)
sum(sr^2)/431
pchisq (sum(sr^2), 431)
```

## Part b :Next extend the model to include pre-treatment measures of the outcome and the additional pre-treatment variables included in the dataset. Does the model fit well? Is there evidence of overdispersion?

The model does not fit well, since the deviance ratio is very large $\frac{10200}{428} = 23.83178 $, though the coefficient are all statistically significant. 

The diviance decreases from 12925 in model 1 to 10200 in model 2 is huge, which means adding other changes the outcome. In treatment group 1 (couple = 1, women_alone = 0), only women of couples participated in the counseling sessions. Couples in this group have about a 41% lower chance than group where both members of couples participated in the session of having unprotected sexual behaviors while other covariates stay the same. 

In treatment group 2 (couples = 0, women_alone = 1), women who participated the study alone have 66% lower chance to have unprotected sexual behavior compared to group where both members of couples participated in the session while other covariates stay the same. 

An increase of unit of number of unprotected sex acts at baseline will decrease the likelihood of having unprotected sexual behavior after the intervention by 1% while other covariates stay the same. 

If the participant is woman, the likelihood of having unprotected sexual behavior after the intervention will increase by 11% while other covariates stay the same. 

And if the baseline HIV status of participant is positive, the likelihood of having unprotected sexual behavior after the intervention will decrease by 43.83% while other covariates stay the same.

```{r}
Mod2 = glm(fupacts ~ treatment + bupacts + factor(sex) + factor(bs_hiv), family=poisson, data=data)
summary(Mod2)
```

And there is overdispersion. The sum of squared standard residuals is  $\sum_{z = 1}^{m}{z_i} = 12841.73$ , compared to an expected value of n - k = 428. The estimated overdispersion factor is $\frac{\sum_{i = 1}^{m}{z_i}}{n-k} = \frac{12841.73}{428} = 30.00404$, and the p-value is 1, indicating that the probability is essentially zero that a random variable from a $\chi^2_{428}$ distribution would be as large as 12841.73. In summary, the risk_behaviors data is overdispersed by a factor of 30, which is huge. 

```{r}
pv1 = fitted(Mod2)
r1= (data$fupacts - fitted(Mod2))

p1_1 = ggplot() +
  geom_point(aes(pv1,r1),colour = "blue", size = 1,show.legend = TRUE) +
  ylab('raw residuals') + 
  xlab('predicted value') +
  geom_hline(yintercept=0, color = "red") +
  theme_bw()

sr1 <- (data$fupacts - fitted(Mod2))/sqrt(fitted(Mod2))
p2_1 = ggplot() +
  geom_point(aes(pv1,sr1),colour = "blue", size = 1,show.legend = TRUE) +
  ylab("standardized residuals") + 
  xlab("predicted value") +
  geom_hline(yintercept=0, color = "red") +
  geom_hline(yintercept=1.96, linetype="dashed", color = "red") +
  geom_hline(yintercept=1.96, linetype="dashed", color = "red") +
  geom_hline(yintercept=-1.96, linetype="dashed", color = "red") +
  theme_bw()
grid.arrange(p1_1,p2_1)

sum(sr1^2)
sum(sr1^2)/428
pchisq(sum(sr1^2), 428)
```

## Part c: Fit an overdispersed Poisson model. What do you conclude regarding effectiveness of the intervention?

We fit an overdispersed model using quassipoisson family to correct for overdispersion. And the correction does not change the coefficient drastically.
The intervention has positive affects on decreasing the unprotected sexual acts.
Group 1 (only women of couples participated in the counseling sessions) has 40.998% less chance to have unprotexted sexual acts than group where both members of couples participated in the session of having unprotected sexual behaviors while other covariates stay the same.

Group2 (women who participated the study alone) has 66.22% less probability to have unprotexted sexual acts than group where both members of couples participated in the session of having unprotected sexual behaviors while other covariates stay the same.


```{r}
data$c.bupacts = (data$bupacts - mean(data$bupacts)) / (2 * sd(data$bupacts))
Mod2.ext <- glm(fupacts ~ treatment + c.bupacts + as.factor(sex) + bs_hiv, family=quasipoisson, data=data)
summary(Mod2.ext)
```
 
## Part d: These data include responses from both men and women from the participating couples. Does this give you any concern with regard to our modeling assumptions?

Yes, because people in the same couple might be highly correlated.

# Q2


# Q3

For part a, the result is pretty the same.
```{r}
library(MASS)
data_mat = model.matrix(fupacts~.,data[,-1])
x = data_mat[,7:8]
x2 = data_mat[,c(2,5,6,7,8)]

y = data$fupacts

betas = c(3,-0.5,-0.5)
betas2 = c(2,-0.4,-0.6,0,0.1,-0.5)

rls = function(x,y,betas,e = 1e-20){
  x = cbind(1,x)
  
  for(i in 1:5000){
  eta = x %*% betas
  mu = exp(eta)
  V = mu
  Z = eta + (y -mu)/mu
  W = as.vector(V/exp(2*eta))
  
  betas_n =lsfit(x,Z,wt = 1/W, intercept = F)$coef
  #betas_n = ginv(t(x) %*% diag(W) %*% x) %*% (t(x) %*% diag(W) %*% Z)
  diff = sum(abs(betas_n - betas))
  
  if(diff >= e) {
    betas = betas_n
    } else 
      {break}
  }
  return(betas_n)
}

rls(x=x, y=y, betas = betas)
rls(x=x2, y=y, betas = betas2)
```
